---
title: "Understanding Bicycle Theft Patterns in Toronto"
subtitle: "A Guide for Cyclists"
author: 
  - John Zhang
  - Yaning Jin
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: true
---
\newpage
# Introduction
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
required_packages <- c(
  "lubridate", "sf", "tidyverse", "bookdown"
  # include any packages to be installed here
) 
for (p in required_packages) { # install if missing
  if (!require(p, character.only = TRUE)) {
    install.packages(p, character.only = TRUE)
  }
}
library(tidyverse)
library(sf)
# you are welcome to use either the CSV file or the GeoJSON file
# GeoJSON format maybe useful if you would like to combine
# an external geospatial dataset
bike_thefts_csv <- read_csv("data/Bicycle_Thefts_Open_Data.csv") |>
  mutate(
    occurence_date = as_date(
      OCC_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    ),
    report_date = as_date(
      REPORT_DATE, format = "%a, %d %b %Y %H:%M:%S GMT"
    )
  )
```
Bicycle theft is a persistent problem in Toronto, posing a considerable deterrent for cyclists who rely on biking as a primary mode of transport. As cycling becomes increasingly popular in the city, understanding theft patterns can help residents make informed choices about their bike security. This study, commissioned by the Cycle Toronto Association, dives deep into Toronto Police Service data on bicycle thefts from 2014 to mid-2024, uncovering patterns related to the timing, location, and context of these incidents. By adjusting for factors like seasonal cycling activity and neighborhood-specific characteristics, we aim to provide insights that will help Toronto cyclists protect their bikes more effectively and encourage safe cycling practices year-round.

Our analysis explores four critical aspects of bike theft risk: geographic hotspots, seasonal timing, theft trends by premises type and time of day, and preferences for specific types of bikes. Each of these areas sheds light on distinct facets of theft risk, which together build a comprehensive picture of when, where, and how bike thefts occur. The findings from this analysis not only serve to inform individual cyclists but also have broader implications for community planning and public awareness efforts. With this background in mind, we first examine where bike thefts are most concentrated in Toronto, starting with a look at the city’s high-theft zones.
\newpage

# Toronto’s High-Theft Zones: Mapping Bicycle Theft Hotspots Across the City
```{r viz-heatmap, fig.cap="Concentrated Bicycle Theft Hotspots in Toronto’s Downtown Core and Waterfront Areas"}
# Load necessary libraries
library(sf)
library(ggplot2)
library(dplyr)

# Load the neighborhood shapefile with suppressed messages using `quiet = TRUE`
neighborhoods <- st_read("data/Neighbourhoods v2_region.shp", quiet = TRUE)

# Check if geometries are valid
validity_check <- st_is_valid(neighborhoods)
if (any(!validity_check)) {
  message("Some geometries are invalid. Applying st_make_valid() to fix them.")
  neighborhoods <- st_make_valid(neighborhoods)
}

# Ensure neighborhoods is in WGS84 (EPSG:4326)
neighborhoods <- st_transform(neighborhoods, crs = 4326)

# Load the bike theft data
bike_thefts <- read.csv("data/Bicycle_Thefts_Open_Data.csv")

# Filter out rows with missing or zero coordinates
bike_thefts <- bike_thefts %>%
  filter(LONG_WGS84 != 0 & LAT_WGS84 != 0)

# Convert the bike theft data to an sf object with WGS84 CRS
bike_thefts_sf <- st_as_sf(bike_thefts, coords = c("LONG_WGS84", "LAT_WGS84"), crs = 4326)

# Perform a spatial join to count bike thefts per neighborhood
theft_counts <- st_join(bike_thefts_sf, neighborhoods, join = st_within) %>%
  group_by(NAME) %>%
  summarise(theft_count = n()) %>%
  st_drop_geometry()  # Drop geometry to simplify the result for merging

# Merge theft counts back with neighborhoods for mapping
neighborhoods <- neighborhoods %>%
  left_join(theft_counts, by = "NAME")

# Replace NA theft counts with zero (for neighborhoods with no recorded thefts)
neighborhoods$theft_count[is.na(neighborhoods$theft_count)] <- 0

# Plot the choropleth map with transparent background
ggplot(data = neighborhoods) +
  geom_sf(aes(fill = theft_count), color = "white") +
  scale_fill_gradient(low = "lightblue", high = "red", name = "Theft Density") +
  theme_void() +
  labs(title = "Bicycle Theft Density by Neighborhood in Toronto") +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5) # Center the title
  )

```

A central question in understanding bike theft risk is geographic: where in Toronto are bikes most likely to be stolen? To answer this, we generated a density map visualizing theft incidents across Toronto neighborhoods. Figure \@ref(fig:viz-heatmap) highlights the downtown core, including the Entertainment District, Kensington Market, and the Waterfront, as the city’s primary theft hotspots. These areas experience the highest density of reported bike thefts, likely due to their status as bustling hubs with high foot traffic and frequent cyclist activity. Toronto’s downtown region is not only dense in population but also features numerous commercial establishments, tourist attractions, and transit hubs, which collectively create a high concentration of bikes parked and unattended at various times of day.

This spatial analysis suggests that the clustering of theft incidents is linked to a combination of factors: the density of cyclists, the convenience of public transit for potential thieves, and the mix of commercial and recreational attractions that draw both residents and visitors. The dense clustering around transit hubs and commercial areas further implies that bike theft risk correlates with locations where monitoring may be challenging due to the volume of people and bikes. This knowledge is valuable for cyclists, who can make safer choices by avoiding parking in isolated or poorly lit areas within these high-theft neighborhoods. For city planners and law enforcement, these insights suggest that targeted interventions, such as improved surveillance and bike-parking facilities in these zones, could help mitigate theft risk. Understanding where thefts happen is a key part of the puzzle. However, to gain a more complete view, we also need to understand when these thefts are most likely to occur. Let's look at seasonal patterns next to see how the risk varies over the course of the year.

\newpage
# Seasonal Risks: Understanding Monthly Theft Peaks and Patterns
```{r viz-bar1, fig.cap="Monthly bicycle theft counts peak in Toronto's top neighborhoods during summer"}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Load the data
bicycle_data <- read.csv('data/Bicycle_Thefts_Open_Data.csv')

# Identify the top 5 neighborhoods with the most thefts and order them
top_neighborhoods <- bicycle_data %>%
  filter(STATUS == "STOLEN") %>%
  count(NEIGHBOURHOOD_158, sort = TRUE) %>%
  slice_head(n = 5)

# Clean up neighborhood names for the legend (remove numbers in brackets)
cleaned_neighborhood_names <- c(
  "Yonge-Bay Corridor",
  "Downtown Yonge East",
  "Wellington Place",
  "St Lawrence-East Bayfront-The Islands",
  "Annex"
)

# Reorder NEIGHBOURHOOD_158 based on theft counts and clean labels
bicycle_data <- bicycle_data %>%
  filter(NEIGHBOURHOOD_158 %in% top_neighborhoods$NEIGHBOURHOOD_158) %>%
  mutate(
    NEIGHBOURHOOD_158 = factor(NEIGHBOURHOOD_158, 
                               levels = top_neighborhoods$NEIGHBOURHOOD_158, 
                               labels = cleaned_neighborhood_names),
    OCC_MONTH1 = factor(OCC_MONTH, levels = c("January", "February", "March", "April", "May", "June", 
                                              "July", "August", "September", "October", "November", "December"))
  )

# Plot theft counts by month for each top neighborhood in ranked order
ggplot(bicycle_data %>% filter(!is.na(OCC_MONTH1)), aes(x = OCC_MONTH1, fill = NEIGHBOURHOOD_158)) +
  geom_bar(stat = "count", position = "dodge") +
  labs(x = "Month", y = "Theft Count", title = "Monthly Bicycle Theft Counts in Top 5 Toronto Neighborhoods", fill = "Neighbourhood") +
  facet_wrap(~ NEIGHBOURHOOD_158, ncol = 3, nrow = 2, scales = "free") +  # Arrange as 3 on top, 2 on bottom, free y-scales
  scale_x_discrete(breaks = c("January", "December")) +  # Show only January and December on x-axis for all facets
  theme_minimal() +  # Use a minimal theme to remove borders
  theme(
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    strip.text = element_blank(),  # Remove facet labels
    legend.position = "right",
    panel.grid.major.x = element_line(color = "grey85", linetype = "solid", size = 0.5),  # Lighter, thinner vertical grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines for a cleaner look
  )


```

```{r viz-bar2, fig.cap="Cycling activity in Toronto peaks in summer, with the highest incidence in July at 81%"}
# Load necessary libraries
library(ggplot2)
library(reshape2)

# Sample data based on the chart, adjust these values with actual data
data <- data.frame(
  Month = factor(c("January", "February", "March", "April", "May", "June", 
                   "July", "August", "September", "October", "November", "December", "Every month of the year"),
                 levels = c("January", "February", "March", "April", "May", "June", 
                            "July", "August", "September", "October", "November", "December", "Every month of the year")),
  All = c(3, 3, 8, 25, 59, 80, 81, 79, 68, 43, 12, 3, 10),
  Utilitarian = c(5, 6, 8, 25, 51, 64, 78, 81, 84, 79, 67, 43, 12),
  Recreational = c(1, 3, 6, 12, 12, 51, 75, 74, 74, 68, 48, 34, 10)
)

# Melt the data for ggplot
data_melted <- melt(data, id.vars = "Month", variable.name = "Type", value.name = "Percentage")

# Filter data to include only "All"
data_melted <- data_melted %>%
  filter(Type == "All") %>%
  mutate(Type = "All (n=5193)")  # Update Type label

# Plot the data
ggplot(data_melted, aes(x = Month, y = Percentage, fill = Type)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_fill_manual(values = c("darkblue")) +
  labs(title = "Monthly Cycling Incidence in Toronto", y = "Percentage", x = NULL) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)  # Center the title
  ) +
  geom_text(aes(label = paste0(Percentage, "%")), 
            position = position_dodge(width = 0.9), vjust = -0.5, size = 2, fontface = "bold") +  # Bold labels
  scale_y_continuous(labels = scales::percent_format(scale = 1), limits = c(0, 100), breaks = seq(0, 100, by = 20))  # Set 100% as max


```

```{r viz-bar3, fig.cap="Relative bicycle theft rates are highest in January and December across Toronto's top theft neighborhoods"}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Load the data
bicycle_data <- read.csv('data/Bicycle_Thefts_Open_Data.csv')

# Define the monthly cycling incidence rates
cycling_incidence <- data.frame(
  OCC_MONTH1 = factor(c("January", "February", "March", "April", "May", "June", 
                        "July", "August", "September", "October", "November", "December"),
                      levels = c("January", "February", "March", "April", "May", "June", 
                                 "July", "August", "September", "October", "November", "December")),
  Cycling_Incidence_Percent = c(3, 3, 8, 25, 59, 80, 81, 79, 68, 43, 12, 3)
)

# Identify the top 5 neighborhoods with the most thefts and order them
top_neighborhoods <- bicycle_data %>%
  filter(STATUS == "STOLEN") %>%
  count(NEIGHBOURHOOD_158, sort = TRUE) %>%
  slice_head(n = 5)

# Clean up neighborhood names for the legend (remove numbers in brackets)
cleaned_neighborhood_names <- c(
  "Yonge-Bay Corridor",
  "Downtown Yonge East",
  "Wellington Place",
  "St Lawrence-East Bayfront-The Islands",
  "Annex"
)

# Filter for top neighborhoods and add cycling incidence rates
bicycle_data <- bicycle_data %>%
  filter(NEIGHBOURHOOD_158 %in% top_neighborhoods$NEIGHBOURHOOD_158) %>%
  mutate(
    NEIGHBOURHOOD_158 = factor(NEIGHBOURHOOD_158, 
                               levels = top_neighborhoods$NEIGHBOURHOOD_158, 
                               labels = cleaned_neighborhood_names),
    OCC_MONTH1 = factor(OCC_MONTH, levels = c("January", "February", "March", "April", "May", "June", 
                                              "July", "August", "September", "October", "November", "December"))
  ) %>%
  left_join(cycling_incidence, by = "OCC_MONTH1")  # Join cycling incidence data

# Calculate monthly theft counts and relative theft rate by neighborhood
relative_theft_data <- bicycle_data %>%
  count(NEIGHBOURHOOD_158, OCC_MONTH1) %>%
  left_join(cycling_incidence, by = "OCC_MONTH1") %>%
  mutate(Relative_Theft_Rate = (n / Cycling_Incidence_Percent) * 100)  # Theft rate per cycling activity

# Plot relative theft rate by month for each top neighborhood with bar plot
ggplot(relative_theft_data, aes(x = OCC_MONTH1, y = Relative_Theft_Rate, fill = NEIGHBOURHOOD_158)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    x = "Month",
    y = "Relative Theft Rate",
    fill = "Neighborhood", title = "Monthly Relative Bicycle Theft Rate in Top 5 Toronto Neighborhoods"
  ) +
  facet_wrap(~ NEIGHBOURHOOD_158, ncol = 3, nrow = 2, scales = "free") +
  scale_x_discrete(breaks = c("January", "December")) +  # Show only January and December on x-axis
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    strip.text = element_blank(),  # Remove facet labels
    legend.position = "right",
    legend.text = element_text(size = 8),   # Reduce legend text size
    legend.title = element_text(size = 9),  # Reduce legend title size
    legend.key.size = unit(0.5, "cm")       # Reduce size of legend keys
  )

```

In Toronto, the risk of bicycle theft varies significantly throughout the year. This seasonal fluctuation makes it essential for cyclists to understand when and where they are most vulnerable to theft. In this section, we examine three complementary visualizations that collectively illustrate the risk of bike theft across different times of the year. By analyzing monthly theft counts, cycling incidence rates, and a calculated metric called the “relative theft rate", we gain a nuanced understanding of when bikes are most at risk.

Figure @ref(fig:viz-bar1) presents the **Monthly Bicycle Theft Counts in the Top 5 Toronto Neighborhoods**—Yonge-Bay Corridor, Downtown Yonge East, Wellington Place, St. Lawrence-East Bayfront-The Islands, and Annex. These neighborhoods consistently report the highest numbers of bike thefts, with peaks generally occurring between May and October. This increase in thefts aligns with warmer weather, when more cyclists take to the streets, increasing bike visibility and availability. While this chart provides valuable insight into overall theft patterns, it does not account for seasonal fluctuations in cycling activity, which can skew the perception of risk. A higher count in the summer months may simply reflect the fact that more people are cycling, not that the per-cyclist risk of theft is necessarily higher.

To address this limitation, we turn to Figure @ref(fig:viz-bar2), a visualization that shows the percentage of cyclists active in each month.^[Source: Nanos Research, online non-probability panel survey, May 13 to 29, 2019, n=5193* Toronto cyclists.] Cycling activity rises sharply in spring, with substantial peaks from May to September, reaching a high of 81% in July. This metric is essential for contextualizing theft data: a higher incidence of theft during these months may simply result from increased cycling activity rather than an elevated risk of theft per cyclist. By understanding cycling incidence, we gain a clearer picture of how theft patterns relate to cyclist behavior and seasonal trends.

Combining these two data points—monthly theft counts and cycling incidence—we derive a more accurate measure of risk: the **Relative Theft Rate**. The relative theft rate is calculated by dividing the number of thefts in each month by the percentage of cyclists active in that month. This adjustment normalizes the data, providing a clearer indication of theft risk per cyclist, independent of seasonal cycling fluctuations. For example, as shown in Figure @ref(fig:viz-bar3), although July reports a high number of thefts, the high cycling incidence dilutes the per-cyclist risk, resulting in a relatively lower relative theft rate than the raw count would suggest. Conversely, winter months like December and January, which see fewer cyclists, show a heightened relative theft rate despite lower overall theft counts. This indicates that cyclists who venture out during these colder months face a higher risk per cyclist, likely due to factors such as limited secure parking and reduced public presence to deter thieves.

The relative theft rate is a more precise measure of individual risk than raw theft counts because it accounts for the seasonal nature of cycling activity. By normalizing the data in this way, we isolate the risk factor from seasonal cycling variations, revealing a clearer picture of individual vulnerability. This insight highlights an essential distinction: although absolute theft counts are higher in the summer, the relative theft risk per cyclist is not necessarily elevated. Instead, winter months may pose a higher per-cyclist risk, signaling a need for additional caution and security measures for those who continue to cycle through colder periods.

Taken together, these visualizations—starting from raw theft counts, moving to cycling incidence rates, and concluding with relative theft rates—offer a comprehensive view of seasonal bike theft risks in Toronto. For cyclists, understanding these patterns can lead to safer practices, such as avoiding high-theft areas during low-cycling months or ensuring better security measures when parking in neighborhoods like Yonge-Bay Corridor, which consistently reports high theft rates year-round. By adopting this data-driven approach, cyclists can make informed decisions that help mitigate the risk of theft throughout the year. While seasonal trends show when thefts are more common, another factor is critical in assessing risk: the context of the theft. Are there certain types of locations or times of day that make bikes more vulnerable? The next section delves into these situational factors.

\newpage
# Context Matters: How Location and Time of Day Affect Bicycle Theft Risk
Our next step was to analyze theft incidents by premises type—such as residential,public locations—and by time of day. This breakdown reveals nuanced differences in theft patterns that are relevant for different kinds of cyclists. For instance, in Figure \@ref(fig:viz-residential) residential areas see distinct patterns based on whether the theft occurs at an apartment or a house. In apartment complexes, thefts peak in the afternoon, likely coinciding with times when bikes are parked but unattended as residents go about their day. Conversely, in house areas, thefts tend to occur in the evening and at night, possibly reflecting a risk associated with overnight storage in more secluded or poorly lit residential spaces.

Public locations, including commercial areas, educational institutions, transit hubs, and outdoor spaces, exhibit their own unique patterns, as shown in Figure \@ref(fig:viz-public). In commercial and outdoor areas, afternoon thefts are most common, aligning with peak business hours when bikes are frequently parked outside shops and public places. Educational institutions, on the other hand, see a morning peak in thefts, which may correspond to the influx of students and staff arriving for the day. Transit hubs display a similar pattern, with higher theft rates in the morning, likely reflecting the periods when cyclists are arriving to use public transportation. These insights provide guidance for cyclists on when and where they should be most cautious. Cyclists parking in commercial areas or at transit hubs should consider choosing high-visibility, secure locations and avoid leaving their bikes unattended for long periods during peak times. This understanding of theft risk by premises type and time of day allows cyclists to take location-specific precautions, which can be particularly valuable for daily commuters or students. Having examined the impact of location and timing, the final piece of the puzzle lies in the bikes themselves. In the following section, we explore which types of bikes are most targeted by thieves and how bike value influences theft risk.

```{r viz-residential, fig.cap="Bicycle thefts peak in the afternoon for apartments and in the evening and night for houses"}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Load the data
data <- read.csv("data/Bicycle_Thefts_Open_Data.csv")

# Extract time of day based on OCC_HOUR (assuming it's in a 24-hour format)
data$Time_of_Day <- cut(
  data$OCC_HOUR,
  breaks = c(-Inf, 6, 12, 18, Inf),
  labels = c("Night", "Morning", "Afternoon", "Evening")
)

# Reorder the factor levels of Time_of_Day for consistency in ordering
data$Time_of_Day <- factor(data$Time_of_Day, levels = c("Morning", "Afternoon", "Evening", "Night"))

# Calculate theft percentages by Premises_Type and Time_of_Day, filtering for "House" and "Apartment"
theft_data_residential <- data %>%
  filter(PRIMARY_OFFENCE == "THEFT UNDER", PREMISES_TYPE %in% c("House", "Apartment")) %>%
  count(PREMISES_TYPE, Time_of_Day) %>%
  group_by(PREMISES_TYPE) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ungroup()

# Define a custom color palette for times of day
custom_colors <- c("Morning" = "#F1C40F", 
                   "Afternoon" = "#3498DB", 
                   "Evening" = "#9B59B6", 
                   "Night" = "#E74C3C")

# Create ring (donut) plots for House and Apartment
ggplot(theft_data_residential, aes(x = 2, y = percentage, fill = Time_of_Day)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  facet_wrap(~ PREMISES_TYPE, ncol = 2) +
  xlim(0.5, 2.5) +  # Create a ring by adjusting x-axis limits
  scale_fill_manual(values = custom_colors) +
  theme_void() +  # Clean up chart appearance
  labs(title = "Bicycle Theft by Time of Day in Residential Areas") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), color = "white", size = 4) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    legend.title = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "bottom",
  )

```

```{r viz-public, fig.cap="Bicycle thefts in public areas peak in the afternoon for commercial, educational, and outside spaces, while transit locations see the highest theft rate in the morning"}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Load the data
data <- read.csv("data/Bicycle_Thefts_Open_Data.csv")

# Extract time of day based on OCC_HOUR (assuming it's in a 24-hour format)
data$Time_of_Day <- cut(
  data$OCC_HOUR,
  breaks = c(-Inf, 6, 12, 18, Inf),
  labels = c("Night", "Morning", "Afternoon", "Evening")
)

# Reorder the factor levels of Time_of_Day for consistent ordering
data$Time_of_Day <- factor(data$Time_of_Day, levels = c("Morning", "Afternoon", "Evening", "Night"))

# Calculate theft percentages by Premises_Type and Time_of_Day
theft_data <- data %>%
  filter(PRIMARY_OFFENCE == "THEFT UNDER") %>%
  count(PREMISES_TYPE, Time_of_Day) %>%
  group_by(PREMISES_TYPE) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ungroup()

# Filter to include only public premises types
public_theft_data <- theft_data %>%
  filter(PREMISES_TYPE %in% c("Commercial", "Educational", "Transit", "Outside"))

# Define a custom color palette for times of day
custom_colors <- c("Morning" = "#F1C40F", 
                   "Afternoon" = "#3498DB", 
                   "Evening" = "#9B59B6", 
                   "Night" = "#E74C3C")

# Create ring (donut) plots for public premises types only
ggplot(public_theft_data, aes(x = 2, y = percentage, fill = Time_of_Day)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  facet_wrap(~ PREMISES_TYPE, ncol = 2) +
  xlim(0.5, 2.5) +  # Create a ring by adjusting x-axis limits
  scale_fill_manual(values = custom_colors) +
  theme_void() +  # Clean up chart appearance
  labs(title = "Bicycle Theft by Time of Day in Public Areas") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), color = "white", size = 2.777) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    legend.title = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "bottom",
  )

```

\newpage
# High-Value Targets: Bike Types and Values That Attract Thieves
Finally in Figure \@ref(fig:viz-stacked) we explored how different types of bikes, along with their value, influence theft risk. Certain types of bikes, such as racing and touring bikes, are more frequently targeted, especially those with values exceeding $1200. This finding underscores the importance of bike value in determining theft risk; thieves appear to prioritize higher-value bikes, which are likely to have a higher resale value. By contrast, less expensive bike types, such as BMX bikes and folding bikes, experience relatively lower theft rates, suggesting they may be perceived as less lucrative targets.

This analysis highlights the need for owners of high-value bikes to invest in enhanced security measures. Such cyclists might consider using GPS tracking devices, heavy-duty locks, or even bike insurance to safeguard against potential theft. Additionally, bike owners who primarily commute might consider choosing a less expensive or more inconspicuous bike if they plan to leave it parked in high-risk areas. For cyclists, understanding the specific vulnerabilities associated with their bike type can help in making decisions around bike purchase and security practices. With this information, they can better weigh the trade-offs between convenience, value, and security. Altogether, these findings highlight various aspects of bicycle theft in Toronto, offering insights that can help cyclists make informed decisions. In the conclusion, we summarize key takeaways and actionable recommendations based on this analysis.

```{r viz-stacked, fig.cap="Higher-value bikes like electric and racing types are more frequently stolen in the 1200+ range, while lower-value bikes are more often stolen in the 0-600 range"}
# Rename columns
data <- data %>%
  rename(BIKETYPE = BIKE_TYPE)

# Proceed with the recode and other processing
data <- data %>%
  mutate(BIKETYPE = recode(BIKETYPE,
                           "BM" = "BMX",
                           "EL" = "ELECTRIC",
                           "FO" = "FOLDING",
                           "MT" = "MOUNTAIN",
                           "OT" = "OTHER",
                           "RC" = "RACER",
                           "RE" = "RECUMBANT",
                           "RG" = "REGULAR",
                           "SC" = "SCOOTER",
                           "TA" = "TANDEM",
                           "TO" = "TOURING",
                           "TR" = "TRICYCLE",
                           "UN" = "UNICYCLE",
                           "UNKNOWN" = "TYPE UNKNOWN"),
         # Define cost categories
         cost_cat = cut(BIKE_COST, breaks = c(0, 600, 1200, Inf), labels = c("0-600", "601-1200", "1200+"), right = FALSE))

# Proceed with the rest of your code
data_cost_premise.p <- data %>% group_by(cost_cat, BIKETYPE) %>% 
  summarise(total = n())
data_cost_premise_d <- data %>% group_by(BIKETYPE) %>% 
  summarise(denom = n()) 
data_cost_premise <- left_join(data_cost_premise.p, data_cost_premise_d, by = "BIKETYPE")
data_cost_premise <- data_cost_premise %>% 
  mutate(proportion = total / denom)

# Plot with swapped axes and angled X-axis text
data_cost_premise %>% 
  filter(!is.na(cost_cat)) %>%
  ggplot(aes(x = proportion, y = BIKETYPE, fill = cost_cat)) +
  geom_bar(stat = "identity", position = position_fill(reverse = TRUE), width = 0.7) +
  geom_text(aes(label = scales::percent_format(accuracy = 1)(proportion)), 
            colour = "white", position = position_fill(vjust = 0.5, reverse = TRUE), size = 3) +
  scale_x_continuous(expand = expansion(0, 0), labels = scales::percent) +  # Convert to percentage
  scale_y_discrete(limits = rev(unique(data_cost_premise.p$BIKETYPE))) + 
  scale_fill_manual(values = c("0-600" = "#F1C40F", "601-1200" = "#3498DB", "1200+" = "#9B59B6"), 
                    guide = guide_legend(title = NULL)) + 
  coord_flip() +
  labs(y = "Bike Type", x = "Proportion") + 
  ggtitle("Distribution of Bike Types by Theft Value Range") +  # Add title here
  theme(legend.position = "top") + 
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 7),    
    axis.text.y = element_text(size = 7),   
    legend.text = element_text(size = 7)     
  )


```

\newpage
# Conclusion
Our analysis has provided several key insights into bicycle theft patterns in Toronto, covering high-risk areas, peak times, vulnerable premises types, and commonly targeted bike types. The findings suggest that Toronto cyclists face higher risks in the downtown core and popular public areas, particularly during the winter months when fewer bikes are on the road. Cyclists should exercise increased caution by securing their bikes in well-lit, monitored locations and considering seasonally adjusted security practices. Furthermore, theft patterns vary significantly by premises type, with distinct risks associated with residential, commercial, and transit locations. Cyclists can leverage this information to avoid high-risk parking practices based on the specific context of where they park their bikes.

Finally, our analysis of bike types underscores the influence of bike value on theft risk. Higher-value bikes are targeted more frequently, suggesting that thieves prioritize expensive models with higher resale potential. By understanding the relationship between bike type, value, and theft likelihood, cyclists can make informed choices about their security needs, whether by investing in stronger locks, choosing more secure parking locations, or even opting for lower-value bikes when commuting.

These insights form a holistic view of bicycle theft risk in Toronto, empowering cyclists to take control of their bike security. While bicycle theft is a complex issue influenced by multiple factors, targeted strategies based on our findings can help reduce individual risk. For the Cycle Toronto Association, these insights may also serve as a foundation for broader public awareness initiatives, helping the city’s cycling community stay safe and secure on Toronto’s streets. As Toronto continues to promote cycling as a sustainable mode of transport, addressing bike theft risks will be essential in building a more bike-friendly city.